A cautionary note on the use of Ornstein Uhlenbeck and other models in macroevolutionary studies

### 2015-05-01 Manuscript/ohyou.tex

% NC: I've moved this up as I think the "uses" section forms part of the intro to the OU model. I'm not convinced uses should come before the model itself but I don't think it matters at this point.
% NC:  This paragraph may need modifying once we finish
  In this paper we present an introduction to the OU model, its general properties and some issues with its use in ecology, evolution and palaeontology.
  We use simulations to demonstrate the inherent bias in estimating the core parameter of the OU model, $\alpha$ that describes the strength of pull towards a central value (typically referred to as the trait or selective optima). We discuss the intricacies of interpreting OU models biologically, and provide advice for appropriate use of OU models in phylogenetic comparative analyses. 
  We also show that very small amounts of intraspecific trait variation (including measurement error) in data can have profound effects on the performance of models. 
  Many of our findings are applicable to other models of evolution (see section X) 
  % Add section no here
  , but we focus on the OU model because of its widespread use and because of the ambiguity in the link between pattern and process when interpreting estimates of the $\alpha$ parameter. 
  We are not the first to describe these problems \citep[e.g.,][]{ho2013asymptotic,ho2014intrinsic,boettiger2012your,hansen2012interpreting,ives2010phylogenetic}, but we hope by summarising the issues and giving recommendations we can help those less familiar with the technical literature.

@@

    \begin{figure}[h]
      \centering
      \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OhYou_Figure1.pdf}
      \caption{The number of ecology, evolutionary biology and palaeontology papers published between 2005 and 2014 containing the phrase ``Ornstein-Uhlenbeck'', as a proportion of the total number of ecology, evolutionary biology or palaeontology papers published that year. See Supporting Information for details. %linkto supp mat
      }
      \label{figure.literature}
    \end{figure}

@@

    \begin{table}[h]
      \begin{center}
        \caption{The most common uses of Ornstein Uhlenbeck models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2013. See Supporting information for details.}
        \bigskip  
          \begin{tabular}{p{8cm}lc}
            \hline
            Use of OU model & Optima & Number of papers\\
            \hline
            Ancestral state reconstruction & single & 8\\
            & multiple & 2\\
            Convergent evolution & single & 0\\
            & multiple & 2\\
            Model of evolution & single & 31\\
            & multiple & 27\\
            Phylogenetic generalised least squares & single & 35\\
            & multiple & 0\\
            Other & single & 5\\
            & multiple & 5\\
            Total & single &  79\\
            & multiple & 36\\
            \hline
            \label{table.uses}
          \end{tabular}
      \end{center}
    \end{table}

@@

      \begin{table}[!htbp]
        \begin{center}
        \caption{Rejection rate and $\alpha$ estimates for data simulated under a constant rate Brownian model on a range of constant-rate birth death trees. Tree type refers to the extinction fraction for the birth-death trees. The rejection rate is the proportion of OU models favoured relative to a Brownian motion model.}
        \bigskip
        \input{Tables/OU_bd_noerror.tex}
        \label{tables:bd_noerror}
        \end{center}
      \end{table}

      \begin{table}[!htbp]
        \begin{center}
        \caption{Rejection rate and $\alpha$ estimates for data simulated under a constant rate Brownian model on trees simulated under time variable speciation rates. The rejection rate is the proportion of OU models favoured relative to a Brownian motion model.}
        \bigskip
        \input{Tables/OU_var_rate_noerror.tex}
        \label{tables:varrate_noerror}
        \end{center}
      \end{table}

@@

    \begin{figure}[h]
      \centering
      \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OhYou_Figure2.pdf}
      \caption{The number of taxa in phylogenies used to fit OU models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2014. Two studies with \textgreater 3000 taxa have been omitted for clarity. See Supporting Information for details.
      }
      \label{figure:ntaxa}
    \end{figure}

    We provide recommendations related to these findings below. % name of section?

@@

    \begin{landscape}
      \begin{table}[!htbp]
        \begin{center}
        \caption{Rejection rate and $\alpha$ estimates for data simulated under a constant rate Brownian model with 0, 1, 5, or 10\% measurement error (m.e.). The rejection rate is the proportion of OU models favoured relative to a Brownian motion model.}
        \bigskip
        \input{Tables/OU_yule_error.tex}
        \label{tables:yule_error}
        \end{center}
      \end{table}
    \end{landscape}

@@

Figure 1:

Figure 2: The number of taxa in phylogenies used to fit OU models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2014. Two studies with \textgreater 3000 taxa have been omitted for clarity. See Supporting Information for details.

>>>>

% NC: I've moved this up as I think the "uses" section forms part of the intro to the OU model. I'm not convinced uses should come before the model itself but I don't think it matters at this point.
% NC:  This paragraph may need modifying once we finish
  In this paper we present an introduction to the OU model, its general properties and some issues with its use in ecology, evolution and palaeontology.
  We use simulations to demonstrate the inherent bias in estimating the core parameter of the OU model, $\alpha$ that describes the strength of pull towards a central value (typically referred to as the trait or selective optima). We discuss the intricacies of interpreting OU models biologically, and provide advice for appropriate use of OU models in phylogenetic comparative analyses. 
  We also show that very small amounts of intraspecific trait variation (including measurement error) in data can have profound effects on the performance of models. 
  Many of our findings are applicable to other models of evolution, but we focus on the OU model because of its widespread use and because of the ambiguity in the link between pattern and process when interpreting estimates of the $\alpha$ parameter. 
  We are not the first to describe these problems \citep[e.g.,][]{ho2013asymptotic,ho2014intrinsic,boettiger2012your,hansen2012interpreting,ives2010phylogenetic}, but we hope by summarising the issues and giving recommendations we can help those less familiar with the technical literature.

@@

    \begin{figure}[!htbp]
      \centering
      \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OhYou_Figure1.pdf}
      \caption{The number of ecology, evolutionary biology and palaeontology papers published between 2005 and 2014 containing the phrase ``Ornstein-Uhlenbeck'', as a proportion of the total number of ecology, evolutionary biology or palaeontology papers published that year. See Supporting Information for details.
      }
      \label{figure.literature}
\end{figure}
% Weird issue with endfloat means these need to have no white space

@@

    \begin{table}[!htbp]
      \begin{center}
        \caption{The most common uses of Ornstein Uhlenbeck models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2013. See Supporting information for details.}
        \bigskip  
          \begin{tabular}{p{8cm}lc}
            \hline
            Use of OU model & Optima & Number of papers\\
            \hline
            Ancestral state reconstruction & single & 8\\
            & multiple & 2\\
            Convergent evolution & single & 0\\
            & multiple & 2\\
            Model of evolution & single & 31\\
            & multiple & 27\\
            Phylogenetic generalised least squares & single & 35\\
            & multiple & 0\\
            Other & single & 5\\
            & multiple & 5\\
            Total & single &  79\\
            & multiple & 36\\
            \hline
            \label{table.uses}
          \end{tabular}
      \end{center}
\end{table}
% Weird issue with endfloat means these need to have no white space

@@

      \begin{table}[!htbp]
        \begin{center}
        \caption{Rejection rate and $\alpha$ estimates for data simulated under a constant rate Brownian model on a range of constant-rate birth death trees. Tree type refers to the extinction fraction for the birth-death trees. The rejection rate is the proportion of OU models favoured relative to a Brownian motion model.}
        \bigskip
        \input{Tables/OU_bd_noerror.tex}
        \label{tables:bd_noerror}
        \end{center}
\end{table} 
% Weird issue with endfloat means these need to have no white space

      \begin{table}[!htbp]
        \begin{center}
        \caption{Rejection rate and $\alpha$ estimates for data simulated under a constant rate Brownian model on trees simulated under time variable speciation rates. The rejection rate is the proportion of OU models favoured relative to a Brownian motion model.}
        \bigskip
        \input{Tables/OU_var_rate_noerror.tex}
        \label{tables:varrate_noerror}
        \end{center}
\end{table} 
% Weird issue with endfloat means these need to have no white space

@@

    \begin{figure}[!htbp]
      \centering
      \includegraphics[width=10cm, height=10cm, keepaspectratio=true]{Figures/OhYou_Figure2.pdf}
      \caption{The number of taxa in phylogenies used to fit OU models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2014. Two studies with \textgreater 3000 taxa have been omitted for clarity. See Supporting Information for details.
      }
      \label{figure:ntaxa}
\end{figure} 
% Weird issue with endfloat means these need to have no white space

    We provide recommendations related to these findings below.

@@

    \begin{landscape}
      \begin{table}[!htbp]
        \begin{center}
        \caption{Rejection rate and $\alpha$ estimates for data simulated under a constant rate Brownian model with 0, 1, 5, or 10\% measurement error (m.e.). The rejection rate is the proportion of OU models favoured relative to a Brownian motion model.}
        \bigskip
        \input{Tables/OU_yule_error.tex}
        \label{tables:yule_error}
        \end{center}
\end{table} 
% Weird issue with endfloat means these need to have no white space
    \end{landscape}

@@

\noindent
\textbf{Figure 1}: The number of ecology, evolutionary biology and palaeontology papers published between 2005 and 2014 containing the phrase ``Ornstein-Uhlenbeck'', as a proportion of the total number of ecology, evolutionary biology or palaeontology papers published that year. See Supporting Information for details.

\noindent
\textbf{Figure 2}: The number of taxa in phylogenies used to fit OU models in ecology, evolutionary biology and palaeontology papers published between 2005 and 2014. Two studies with \textgreater 3000 taxa have been omitted for clarity. See Supporting Information for details.
